Leaf mass per area, not total leaf area, drives differences in above-ground biomass distribution among woody plant functional types

### 2016-04-26 manuscript.Rnw
Missing reference and minor tidying

% Notes:
% for submission, convert to Word docx using Adobe Acrobat Professional. To not include line numbers here, add them in Word. Also do not include figures. Word count can appear on first page; follow instructions further below.

@@

% Flag to include figures in the main text
\newif\ifincludefigs
\includefigstrue  % or \includefigsfalse

@@

% Word count
%% Uncomment this after compiling the ms, then compile again.
% Requires Perl to be installed.
%' <<echo=FALSE>>=
%' #shell("texcount manuscript.tex > wordcount.txt")
%' nrw <- get_wordcount("wordcount.txt")
%' @
%' Total number of words: \Sexpr{nrw$Total} \\
%' Abstract: \Sexpr{nrw$Abstract} \\
%' Introduction: \Sexpr{nrw$Introduction} \\
%' Methods and Materials: \Sexpr{nrw$Methods} \\
%' Results: \Sexpr{nrw$Results} \\
%' Discussion: \Sexpr{nrw$Discussion} \\
%' Number of figures: 6 (all in colour) \\
%' Number of tables: 2 \\
%' Supporting information : 6 Figures (all in colour).

% Alternatively, in a shell,
%
%texcount manuscript.tex > wordcount.txt
%shell("texcount manuscript.tex > wordcount.txt")

@@

In our dataset, stem cross-sectional area was measured either at breast height (1.3 - 1.34m) and/or at the base of the plant. In our analyses, we used basal stem area because many plants (N=\Sexpr{nrow(dataset[dataset$h.t < 1.3,])}) were \textless{} 1.3m tall. We therefore converted records where area was measured at breast height only to area at base, using a relationship fitted to individuals where both measurements existed.  For the subset of the data where both areas were measured, we estimated basal stem area ($\as$) from breast height stem area ($\asb$) from the equation,
\begin{equation}
\label{eq:dia}
D_{\rm S} = D_{\rm Sbh} \left(\frac{H}{H - H_{\rm bh}}\right)^c,
\end{equation}
where $D_{\rm S}$ is the basal stem diameter (m), $D_{\rm Sbh}$ is the stem diameter at breast height, and $H_{bh}$ is the height at which $D_{\rm Sbh}$ was measured (typically 1.3 or 1.34m). Stem area and diameter were always related assuming the stem was circular. We chose to estimate missing stem diameter rather than stem area because a much better fit was obtained, with a more constant error variance. The parameter $c$ in eq. \ref{eq:dia} was further expressed as a function of plant height:
\begin{equation}
c = c_0H^{c_1}.
\end{equation}
The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}, R$^2$ = \Sexpr{signif(nlsr2(basalafit),3)}. The relationship was fit using a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs.

@@

All analyses were conducted in R 3.2.4 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, and `remake` \citep{fitzjohn2015}. In addition, the code replicating this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{will be added at proof stage}).

@@

These patterns in biomass distribution reflected differences in the distribution of components of Eqn \ref{eq:mfmspart}-\ref{eq:afmspart}. In particular, LMA differed substantially among PFTs, whereas $\afas$ did not (Fig. 4), which is consistent with the large differences between PFTs in LMA shown in Fig. 3b.  When PFT was added to the statistical model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfaf$ but not $\afas$  (Table \ref{table:varparttable}), demonstrating that PFT differences entered largely in leaf mass-based variables.

@@

Climate variables consistently added little explanatory power when added to a statistical model that included PFT and height (the adjusted R$^2$ frequently decreased, see Table \ref{table:varparttable}). The only exception was LMA, for which a model that included MAT explained 53\% of the variation, compared to a model with height and PFT (45\%). Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citet{reich2014}, we also analysed climate effects with methods similar to that study. When we tested for MAT effects on $\mfms$ when only considering gymnosperms and angiosperms (thus grouping deciduous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig. S4) for angiosperms for most size classes (but never for gymnosperms). We found, however, that this was largely due to the fact that at higher MAT, an increasing proportion of angiosperms are evergreen (Fig. 1), which have higher $\mfms$. When analysed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig. S5).

@@

Although we found no consistent effects of climate on biomass distribution after PFT and plant height were accounted for, effects of climate may also arise by altering the average height of a stand. At a global scale, average plant height is clearly related to climate via water availability \citep{klein2015, simard2011}. Since biomass distribution itself is steeply related to plant height (Figs. 2, 3), we expect correlations between climate variables and observed biomass distribution when plant height is not first accounted for. Our database did not lend itself well to a direct test of this hypothesis, due to uneven sampling across the range in plant height along climate gradients.

>>>>

% Notes:
% - For submission, convert to Word docx using Latex2rtf. 
% - Do not include line numbers here, add them in Word. Also do not embed figures. 
% - Word count can be automatically added on first page; follow instructions further below.

@@

% Flag to include figures in the main text
\newif\ifincludefigs
\includefigsfalse % or \includefigstrue

@@

% -- Word count
% After compiling the ms (and no cleanup, so .tex is present), run:
% shell("texcount manuscript.tex > wordcount.txt")
% Then uncomment the following:
%' <<echo=FALSE, eval=FALSE>>=
%' nrw <- get_wordcount("wordcount.txt")
%' @
%' Total number of words: \Sexpr{nrw$Total} \\
%' Abstract: \Sexpr{nrw$Abstract} \\
%' Introduction: \Sexpr{nrw$Introduction} \\
%' Methods and Materials: \Sexpr{nrw$Methods} \\
%' Results: \Sexpr{nrw$Results} \\
%' Discussion: \Sexpr{nrw$Discussion} \\
%' Number of figures: 6 (all in colour) \\
%' Number of tables: 2 \\
%' Supporting information : 6 Figures (all in colour).

@@

In our dataset, stem cross-sectional area was measured either at breast height (1.3 - 1.34m) and/or at the base of the plant. In our analyses, we used basal stem area because many plants (N=\Sexpr{nrow(dataset[dataset$h.t < 1.3,])}) were \textless{} 1.3m tall. We therefore converted records where area was measured at breast height only to area at base, using a relationship fitted to individuals where both measurements existed.  For the subset of the data where both areas were measured, we estimated basal stem area ($\as$) from breast height stem area ($\asb$) from the equation,
\begin{equation}
\label{eq:dia}
D_{\rm S} = D_{\rm Sbh} \left(\frac{H}{H - H_{\rm bh}}\right)^c,
\end{equation}
where $D_{\rm S}$ is the basal stem diameter (m), $D_{\rm Sbh}$ is the stem diameter at breast height, and $H_{bh}$ is the height at which $D_{\rm Sbh}$ was measured (typically 1.3 or 1.34m). Stem area and diameter were always related assuming the stem was circular. We chose to estimate missing stem diameter rather than stem area because a much better fit was obtained, with a more constant error variance. The parameter $c$ in Eqn \ref{eq:dia} was further expressed as a function of plant height:
\begin{equation}
c = c_0H^{c_1}.
\end{equation}
The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}, R$^2$ = \Sexpr{signif(nlsr2(basalafit),3)}. The relationship was fit using a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs.

@@

All analyses were conducted in R 3.2.4 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, and `remake` \citep{fitzjohn2015}. In addition, the code replicating this analysis (and all figures) is available at http://github.com/RemkoDuursma/baadanalysis (doi: \emph{will be added at proof stage}).

@@

These patterns in biomass distribution reflected differences in the distribution of components of Eqn \ref{eq:mfmspart}-\ref{eq:afmspart}. In particular, LMA differed substantially among PFTs, whereas $\afas$ did not (Fig. 4), which is consistent with the large differences between PFTs in LMA shown in Fig. 3b, and the substantial variance in LMA explained by just PFT (Table \ref{table:varparttable}).  When PFT was added to the statistical model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfaf$ but not $\afas$  (Table \ref{table:varparttable}), demonstrating that PFT differences entered largely in leaf mass-based variables.

@@

Climate variables consistently added little explanatory power when added to a statistical model that included PFT and height (the adjusted R$^2$ frequently decreased), or to a model that included only PFT (see Table \ref{table:varparttable}). The only exception was LMA, for which a model that included MAT, PFT and height explained 53\% of the variation, compared to a model with height and PFT (45\%). Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citet{reich2014}, we also analysed climate effects with methods similar to that study. When we tested for MAT effects on $\mfms$ when only considering gymnosperms and angiosperms (thus grouping deciduous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig. S4) for angiosperms for most size classes (but never for gymnosperms). We found, however, that this was largely due to the fact that at higher MAT, an increasing proportion of angiosperms are evergreen (Fig. 1), which have higher $\mfms$. When analysed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig. S5).

@@

Although we found no consistent effects of climate on biomass distribution after PFT and plant height were accounted for, effects of climate may also arise by altering the average height of a stand. At a global scale, average plant height is clearly related to climate via water availability \citep{klein2015, simard2011}. Since biomass distribution itself is steeply related to plant height (Figs. 2, 3), we expect correlations between climate variables and observed biomass distribution when plant height is not first accounted for. However, our database did not lend itself well to a direct test of this hypothesis, and indeed we did not find any correlations between any of the studied variables and climate when height was not first accounted for (Table 2).
